GWAS datasets for FDR benchmarking

### 2017-09-21 datasets/GWAS/GWAS.Rmd
Add demo evaluation on 10K SNPs

---
title: "GWAS datasets for FDR benchmarking"
author: "Keegan Korthauer"
date: "9/12/2017"
output: html_document
---

@@

We'll also run the benchmark methods on each dataset and report the number of 
discoveries for each one.

@@

```{r, wkspace-setup}
# Note that swfdr package depends on R version 3.4 or higher
setwd("/n/irizarryfs01_backed_up/kkorthauer/benchmark-fdr/")
source("simulation_studies/R/simulation-helpers.R")
source("datasets/R/evaluation-helpers.R")
setwd("./datasets/GWAS")

@@

```{r, GWAS1-dat}
bmi <- bmi %>% 
        rename(pval = p_value,
               SE = stderr,
               effect_size = effect,
               ind_covariate = N)
```

@@

```{r, GWAS1-run}
cutoffs <- c(0.01, 0.025, 0.05, 0.10, 0.20)
if (!file.exists("bmi_results.RData")){
  bmi_results <- run_benchmarks(dat=bmi, alphas=cutoffs)
  save(bmi_results, file="bmi_results.RData")
}else{
  load("bmi_results.RData")
}

# plot
ggplot(bmi, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("BMI GWAS Metanalysis with Sample Size Covariate")
```

@@

```{r, GWAS2-dat}
asthma <- asthma %>% 
        rename(pval = P_fix,
               SE = se_theta_fix,
               effect_size = theta_fix,
               ind_covariate = no_studies_meta)
```

@@

```{r, GWAS2-run}
# use same cutoffs as for the bmi study (defined above)
if (!file.exists("asthma_results.RData")){
  asthma_results <- run_benchmarks(dat=asthma, alphas=cutoffs)
  save(asthma_results, file="asthma_results.RData")
}else{
  load("asthma_results.RData")
}

# plot
ggplot(asthma_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Asthma GWAS Metanalysis with Sample Size Covariate")
```

@@

```{r, GWAS3-dat}
trig <- trig %>% 
        rename(pval = P_value,
               SE = SE,
               effect_size = beta,
               ind_covariate = N)
```

@@

```{r, GWAS2-run}
# use same cutoffs as for the bmi study (defined above)
if (!file.exists("triglyceride_results.RData")){
  trig_results <- run_benchmarks(dat=trig, alphas=cutoffs)
  save(trig_results, file="triglyceride_results.RData")
}else{
  load("triglyceride_results.RData")
}

# plot
ggplot(trig_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Asthma GWAS Metanalysis with Sample Size Covariate")
```

>>>>

---
title: "GWAS datasets for FDR benchmarking"
author: "Keegan Korthauer"
date: "9/21/2017"
output: html_document
---

@@

We'll also run the benchmark methods on each dataset and report the number of 
discoveries for each one. At first, we'll demo on just a __**small subset of 
SNPs**__, since the computation is time-intensive. After that is working 
smoothly, we'll scale it up to the full set of SNPs.

@@

```{r, wkspace-setup}
# Note that swfdr package depends on R version 3.4 or higher
setwd("/n/irizarryfs01_backed_up/kkorthauer/FDR")
source("benchmark-fdr/simulation_studies/R/simulation-helpers.R")
source("benchmark-fdr/datasets/R/evaluation-helpers.R")
setwd("DATA/GWAS")

@@

```{r, GWAS1-dat}
bmi <- bmi %>% 
        rename(pval = p_value,
               SE = stderr,
               effect_size = effect,
               ind_covariate = N) %>%
        mutate(zscore = effect_size/SE)

# Demo on a small subset of SNPs first (remove this later)
bmi <- bmi[sample(1:nrow(bmi), 10000),]
```

@@

```{r, GWAS1-run, results="hide"}
cutoffs <- c(0.01, 0.025, 0.05, 0.10, 0.20)
resfile <- paste0("bmi_results_", nrow(bmi), ".RData")
if (!file.exists(resfile)){
  bmi_results <- run_benchmarks(dat=bmi, alphas=cutoffs)
  save(bmi_results, file=resfile)
}else{
  load(resfile)
}

# plot N rejects vs FDR cutoff
p <- ggplot(bmi_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("BMI GWAS Metanalysis with Sample Size Covariate") +
  theme_classic()
print(p)

# log-scale y axis
print(p + scale_y_continuous(trans="log2") + 
        ylab("Number of rejections (log2 scale)"))
```

@@

```{r, GWAS2-dat}
asthma <- asthma %>% 
        rename(pval = P_fix,
               SE = se_theta_fix,
               effect_size = theta_fix,
               ind_covariate = no_studies_meta) %>%
        mutate(zscore = effect_size/SE)

# Demo on a small subset of SNPs first (remove this later)
asthma <- asthma[sample(1:nrow(asthma), 10000),]
```

@@

```{r, GWAS2-run, results="hide"}
# use same cutoffs as for the bmi study (defined above)
resfile <- paste0("asthma_results_", nrow(asthma), ".RData")
if (!file.exists(resfile)){
  asthma_results <- run_benchmarks(dat=asthma, alphas=cutoffs)
  save(asthma_results, file=resfile)
}else{
  load(resfile)
}

# plot N rejects vs FDR cutoff
p <- ggplot(asthma_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Asthma GWAS Metanalysis with Sample Size Covariate") +
  theme_classic()
print(p)

# log-scale y axis
print(p + scale_y_continuous(trans="log2") + 
        ylab("Number of rejections (log2 scale)"))
```

@@

Note that for this dataset, all reported effect sizes are positive. This is because
the alleles (A1 and A2) have been ordered such that the direction of effect are 
consistent. Since Scott assumes that the Z-scores are standard normal, we can't 
apply this method to a set of all positive Z-scores. To recover the direction of 
effect, we use the `Freq.A1.1000G.EUR` (allele frequency of A1) variable,
and use it to determine whether
`A1` (allele 1) is the reference allele or not. If the allele frequence of A1 is
greater than 0.5, we consider it the reference allele. Then we'll use the 
comparison of reference to minor as the positive direction. That is,
if A2 was reference,
then effect direction is negative. Otherwise it is positive. Since some of the
SNPs are missing the reference allele information, we cannot determine their 
direction of effect and they will be removed.

```{r, GWAS3-dat}
trig <- trig %>% 
        rename(pval = "P-value",
               SE = se,
               effect_size = beta,
               ind_covariate = N)%>%
        mutate(zscore = effect_size/SE)

# add effect size
trig <- trig %>% 
       mutate(A2.reference = Freq.A1.1000G.EUR < 0.5) %>%
       mutate(zscore = zscore * (1-2*A2.reference))

# remove SNPs with missing values for zscore (couldn't determine direction of 
# effect)
trig <- trig %>% 
  filter(!is.na(zscore))

# Demo on a small subset of SNPs first (remove this later)
trig <- trig[sample(1:nrow(trig), 10000),]
```

@@

```{r, GWAS3-run, results="hide"}
# use same cutoffs as for the bmi study (defined above)
resfile <- paste0("trig_results_", nrow(trig), ".RData")
if (!file.exists(resfile)){
  trig_results <- run_benchmarks(dat=trig, alphas=cutoffs)
  save(trig_results, file=resfile)
}else{
  load(resfile)
}

# plot N rejects vs FDR cutoff
p <- ggplot(trig_results, aes(alpha, n_rejects, color=method)) +
  geom_point() +
  geom_line() +
  xlab("FDR cutoff") + ylab("Number of rejections") +
  ggtitle("Triglyceride GWAS Metanalysis with Sample Size Covariate") +
  theme_classic()
print(p)

# log-scale y axis
print(p + scale_y_continuous(trans="log2") + 
        ylab("Number of rejections (log2 scale)"))

```
